The role of advection versus diffusion in intraventricular transport

### 2025-02-04 paper/main.tex
clean, draft conclusion, prepare comments

\documentclass[fleqn]{wlscirep}
\usepackage[utf8]{inputenc}
\usepackage{lipsum}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsfonts}
\usepackage{siunitx}
\usepackage{physics}
\usepackage{stmaryrd} % double brackets for jump operator
\usepackage{nameref}
\usepackage{booktabs}
\usepackage{bm}
\usepackage{amsopn}
\usepackage[skip=10pt]{caption}
\usepackage{color}
\usepackage{xcolor,colortbl}
\usepackage[nameinlink, capitalise]{cleveref}
\usepackage{graphicx}
\usepackage{tkz-euclide} % draw tikz pictures
\usepackage{subcaption}
\usepackage{tablefootnote} % for table footnotes
\usepackage{float} % figure placement

@@

% Title
\title{The role of advection versus diffusion in intraventricular transport}
\author[1,*]{Halvor Herlyng}
\author[1]{Ada J.~Ellingsrud}
\author[1]{Miroslav Kuchta} 
\author[X]{Inyoung Jeong} % OK?
\author[1,3,+]{Marie E.~Rognes} % OK order?
\author[2,+, *]{Nathalie Jurisch-Yaksi}
\affil[1]{Department of Numerical Analysis and Scientific Computing, Simula Research Laboratory, Oslo, Norway}
\affil[2]{Department of Clinical and Molecular Medicine, Norwegian University of Science and Technology, Trondheim, Norway}
\affil[3]{K.~G.~Jebsen Center for Brain Fluid Research}
\affil[*]{\href{mailto:hherlyng@simula.no}{hherlyng@simula.no} \href{mailto:nathalie.jurisch-yaksi@ntnu.no}{nathalie.jurisch-yaksi@ntnu.no}}
\affil[+]{these authors contributed equally to this work}
% Optional PDF information
\ifpdf
\hypersetup{
  pdftitle={zebrafish-cilia-csf},
  pdfauthor={H.~Herlyng~\emph{et al.}}
}
\fi

@@

\section*{Methods}
\subsection*{Image-based computational geometries of zebrafish brain ventricles}
To model flow and transport in zebrafish larval brain ventricles, we generated a
3D representation of the brain ventricular walls using confocal imaging data of a zebrafish embryo
injected intraventricularly with a 70 kDa dye at 2 dpf~\cite{Olstad2019CiliaryDevelopment}. 
From the imaging data, we constructed a computational mesh of the ventricles (\Cref{fig:fig1}d).
The interior volume was meshed with fTetWild~\cite{Hu2020FastWild}. 
The \emph{standard} mesh consisted of 132 134 tetrahedral cells and had a
maximal (minimal) edge length of 14.9 $\mu$m (4.68 $\mu$m). 
At 2 dpf, the ventricular system consists mainly of 3 cavities: the telencephalic (anterior),
the di-/mesencephalic (middle) and the rhombencephalic (posterior) ventricles, which are connected by ducts. 
Parts of the ventricular surfaces were marked to be lined with
motile cilia~\cite{Olstad2019CiliaryDevelopment} (\Cref{fig:fig1}d, magenta markers).
\begin{figure}%[hbt!]
    \centering
    \includegraphics[width=\linewidth]{graphics/figure1.png}
    \caption{\fixme{Reorder panels.} 
    \textbf{a.} Schematic illustration of the two cerebrospinal fluid (CSF) flow components modeled,
    motile cilia and cardiac pulsations, and their contributions to the flow patterns.
    \textbf{b.} Particle tracking data visualizing flow directionality in the middle ventricle.
    \textbf{c.} Flow velocity magnitudes in the middle ventricle from particle tracking data.
    \textbf{d.} Computational mesh with marked facets indicating regions of motile cilia (magenta) and
    anterior/posterior facets (blue).
    \textbf{e.} Streamlines of the (steady-state) CSF flow field simulated with the
    cilia-only/no-cardiac flow model.
    \textbf{f.} Streamlines of the CSF flow field simulated with the cardiac-only/no-cilia flow
    model at a time 25\% into the cardiac cycle. %t=0.112613 s, period=0.45045 s
    \textbf{g.} Streamlines simulated with the baseline flow model at
    a time 50\% into the cardiac cycle. %t=0.225225 s
    \textbf{h.} Streamlines simulated with the baseline flow model at
    a time 70\% into the cardiac cycle. %t=0.315315 s
    }
    \label{fig:fig1}
\end{figure}

@@

\subsection*{Computational CSF dynamics}
The beating motion of the cilia generates CSF flow with steady rotational
structures~\cite{Olstad2019CiliaryDevelopment}
within the brain ventricles (\Cref{fig:fig1}a--c).
In addition, cardiac pulsations induces pulsatile CSF flow.
We model this flow of CSF in the brain ventricles by the time-dependent,
incompressible Stokes equations which read as follows.
Find the CSF velocity $\uu = \uu(\xx, t) = (u_x(\xx, t), u_y(\xx, t), u_z(\xx, t))$
and the CSF pressure $p = p(\xx, t)$ for $\xx \in \Omega$ and time $t>0$, such that
\begin{subequations}
    \begin{align}
      \rho \pdifft{\uu} - \nabla \cdot \bsig &= \mathbf{0}
      \quad \text{in } \Omega,\\
      \nabla \cdot \uu &= 0
      \quad  \text{in } \Omega,
  \end{align}
  \label{eq:stokes_eqs}% 
\end{subequations}%
where the stress tensor is defined as $\bsig(\uu, p) = 2\mu\beps(\uu) - p\mathbf{I}$
and $\beps(\uu) = \frac{1}{2}\left(\nabla \uu + (\nabla\uu)^T\right)$ is the strain-rate tensor. 
Bold-face characters denote vectors or tensors, and $\mathbf{I}$ is the identity tensor in three dimensions.
Furthermore, $\rho=1000 \ \mathrm{kg/m^3}$ is the CSF density
and $\mu=6.97\times 10^{-4} \ \mathrm{Pa\cdot s}$ is the dynamic viscosity
of the CSF~\cite{Bloomfield1998EffectsFluid}.
We introduce the tangential traction $\bsigpar$ as the
tangential component of the traction $\hat{\bsig}=\bsig\nn$:
\begin{equation*}
    \bsigpar = P_{\nn}(\hat{\bsig}) = (\mathbf{I} - \nn\otimes\nn)\hat{\bsig},
\end{equation*}
where $\nn$ is the outer unit normal of the boundary surface $\Gamma = \partial\Omega$,
and $P_{\nn}(\rr)$ is the tangential projection of a vector $\rr$ onto $\Gamma$.
We consider three types of boundary conditions:
(i) \emph{tangential traction} (no flow normal to the boundary
$\uu \cdot \nn = 0$ with a force $\btau$ applied tangentially on the boundary $\bsigpar = \btau$),
(ii) \emph{normal pressure} (a pressure $\Tilde{p}$ applied in
the normal direction: $(\mu\nabla\uu - p\mathbf{I})\nn = \Tilde{p}(t)\nn$),
and (iii) \emph{free slip} (no flow normal to the boundary
$\uu \cdot \nn = 0$ and no tangential forces $\bsigpar = {\bf 0}$).
Unless noted otherwise, the free-slip condition is applied on the
boundary. We assume that the system starts at rest: $\uu(\xx, t=0) = \mathbf{0}$.

@@

\subsection*{Cilia-driven flow}
On the ciliated regions of the ventricle walls $\Gc$ (\Cref{fig:fig1}d, magenta markers),
we impose a constant-in-time tangential traction $\bsigpar=\btau$ to represent the net forces
of the cilia acting on the CSF.
We set $\btau(\xx, \rr) = \tau \lambda(\xx) P_{\nn}(\rr)$, where the function $\lambda(\xx)$
and the sign of the vector $\rr=\pm(1, 0, 1)$ varies depending on the
cilia region (see~\Cref{subsec:appendixA1}).
Both $\rr$ and the value $\tau=6.5\times 10^{-4}$ Pa were chosen by 
model calibration using experimental data~\cite{Olstad2019CiliaryDevelopment}.
The data indicate speeds proximate to the ventricular walls of $27.4 \pm 5.4$ \textmu m/s and
$2.6 \pm 0.6$ \textmu m/s in the dorsal and ventral regions of the middle ventricle, respectively.
In the anterior ventricle, the speeds were estimated at $4.7 \pm 1.3$ \textmu m/s.
We calibrated $\tau$ heuristically by comparing these experimental velocity values
with the simulated, maximum velocity magnitude.

\subsection*{Flow induced by cardiac pulsations}\label{subsec:pressure_bcs}
To drive a cardiac-induced pulsatile flow inside the brain ventricles, we set $\Tilde{p}(t)=0$ and
$\Tilde{p}(t)=-A\sin{\omega t}$ on anterior and posterior parts, respectively, of the
boundary $\Gp$~(\Cref{fig:fig1}d, blue markers).
Based on a measured cardiac frequency of $f=2.22$ Hz~\cite{Olstad2019CiliaryDevelopment},
the angular cardiac frequency $\omega=2\pi f=6.97$ rad/s.
The amplitude $A=1.5\times 10^{-3}$ Pa determines the magnitude of the pulsatile flow,
and was calibrated based on experimental data~\cite{Olstad2019CiliaryDevelopment}.
Similar to $\tau$, the amplitude $A$ was chosen by comparing experimental and simulated velocities.
Particle tracking data of pulsatile motion was used to determine the contributions of the
cardiac pulsations to the velocity magnitude,
as compared to the effect of cilia motion on the flow (see~\Cref{sec:appC_particle_transport}).

@@

\subsection*{Intraventricular injection of microbeads and particle tracking}
To recover the flow fields shown in~\Cref{fig:fig1}b, c, we carried out
particle tracking in a larval zebrafish injected with fluorescent beads.
As described in Olstad~\emph{et al.}~\cite{Olstad2019CiliaryDevelopment},
anesthetized 2 dpf zebrafish were injected with 1 nl of a mixture containing 0.1\% w/v fluorescent beads
(SPHERO Fluorescent Yellow Particles 1\% w/v, F = 0.16 mm) diluted
in 7.5 mg/ml 70 kDa rhodamine B isothiocyanate-dextran (RITCdextran; Sigma-Aldrich, R9379)
dissolved in artificial CSF. Artificial CSF composition was as follows:
124 mM NaCl, 22 mM D-(+)-Glucose, 2.0 mM KCl, 1.6 mM MgSO\textsubscript{4} $\cdot$ 7 H\textsubscript{2}O,
1.3 mM KH\textsubscript{2}PO\textsubscript{4}, 24 mM NaHCO\textsubscript{3},
2.0 mM CaCl\textsubscript{2} $\cdot$ 2 H\textsubscript{2}O.
The needles used for the injections were pulled with a
Sutter Instrument Co. Model P-2000 from thin-walled glass
capillaries (1.00 mm; VWR) and cut open with a forceps.
A volume of 1 nl of the solution was injected with a pressure injector
(Eppendorf Femtojet 4i) in the rostral rhombencephalic ventricle.
The pressure and time were calibrated for each needle using a 0.01 mm calibration slide.

Following injection, the zebrafish was directly placed under the confocal
microscope and 1200 images were acquired at a frequency of 13.1 Hz.
A single optical section was obtained. Particle tracking was done using
TrackMate~\cite{Tinevez2017TrackMate:Tracking} in Fiji/ImageJ~\cite{Schindelin2012Fiji:Analysis}
and plotted using MATLAB, as described in Jeong~\emph{et al.}~\cite{Jeong2022MeasurementTelencephalon}.
The parameters for TrackMate were as follows.
The DoG detector was used with the settings: threshold: 2.0, with
median filtering; radius: 2.0, with subpixel localization.
Next, the Simple LAP tracker was used with settings: max frame gap: 2;
alternative linking cost factor: 1.05; linking max distance: 5.0; gap closing max distance: 2.0;
splitting max distance: 15.0; allow gap closing: true; allow track splitting:
false; allow track merging: false; merging max distance: 15.0; cutoff percentile: 0.9.
Only the particle tracks with at least 30 data points, while simultaneously
covering a distance of at least 8.8 $\mu$m, were plotted in~\Cref{fig:fig1}b, c.

@@

We seek to model the compartimentalized and pulsatile CSF flow observed \emph{in vivo}~(\Cref{fig:fig1}b--c).
To characterize CSF flow patterns predicted by
computational fluid dynamics within the brain ventricles, we begin by
considering the simulated CSF velocity and pressure when only considering
cilia motion, disregarding cardiac pressure pulsations. The
bidirectional traction induced by the motile cilia
inherently leads to a partial compartmentalization of the ventricular
system, with large-scale vortex structures in the anterior and middle
ventricles~(\Cref{fig:fig1}e). The highest velocities ($26.5$ \textmu m/s)
occur in the vicinity of the dorsal cilia in the middle ventricle.

To examine the effect of cardiac pressure pulsations,
we next consider a no-cilia model variation. In this scenario, the simulated CSF
flow is pulsatile as a result of the sinusoidal pressure gradient (\Cref{fig:fig1}g),
with a peak velocity magnitude of $4.8$ \textmu m/s. At each time instant, the flow
is unidirectional, with CSF initially flowing rostrocaudally (positive $x-$axis) before changing
direction to caudorostral in the middle of the cardiac cycle.

We now consider the baseline model, simulating flow driven by both cilia
and cardiac pulsations. We observe persistent flow features,
with vortex-structures in the anterior and middle ventricles
and directional flow in the posterior ventricle (\Cref{fig:fig1}e,
\fixme{Supplementary video 1}). These characteristic patterns suggest
that the flow in the anterior and middle ventricles are dominated by
the cilia contribution, while the cardiac pressure pulsations dominate
within the posterior ventricle. Over one cardiac cycle, the CSF flow
speed reaches up to 27.7 \textmu m/s, dorsally in the posterior part
of the middle ventricle. This peak is attained at the time instant
coinciding with the time of peak caudal cardiac pulsatile flow, as a
result of both the traction induced by the cilia and the pulsatile
pressure-driven flow being aligned with the rostrocaudal 
axis. Using the observed maximum velocity magnitude and choosing the height of
the middle ventricle ($110$ \textmu m) as a characteristic length
scale, yields a Reynolds number $\mathrm{Re} \approx 0.004$,
justifying the assumption of Stokes flow. In summary, the computational
CSF flow model reproduces the flow features observed \emph{in vivo}.

@@

\subsection*{Absence of ciliary motion affects local protein distribution}
Since CSF flow is highly disturbed in the absence of motile cilia, we
want to assess the impact of cilia motion on solute distribution. To this end,
we measured the movement of photoconverted Dendra2 signal
in control and motile-cilia mutant zebrafish (\emph{schmalhans} (\emph{smh})).
By plotting the means of $\Delta F$ (\Cref{fig:figure4_compare_control_mutant}a)
and the time to threshold
across the control and mutant cohorts (\Cref{fig:figure4_compare_control_mutant}b),
we identified a significant
difference in Dendra2 distribution in ROIs 2 and 4, but not in ROIs 3, 5, and 6. In ROIs 2 and 4,
which measure Dendra2 distribution downstream of the photoconversion site,
we observed a slower increase of fluorescence intensity,
resulting in $\sim X\%$ (ROI 2) and $\sim 45\%$ (ROI 4)
higher mean times-to-threshold in the \emph{smh}
mutants compared to controls (\Cref{fig:figure4_compare_control_mutant}b).

@@

We identified that the velocities were most strongly
affected when removing cilia in the dorsal diencephalic cilia,
as the maximum velocity dropped from 27.7 $\mu$m/s to 8.0 $\mu$m/s.
\emph{In vivo}, these cilia generate the strongest flow at
this developmental stage of the fish~\cite{Olstad2019CiliaryDevelopment}.
In contrast, removing telencephalic and ventral diencephalic had minimal
impact and resulted in maximum velocities of 27.7 $\mu$m/s or 27.2 $\mu$m/s, respectively.
Changes in the transport dynamics of Dendra2 were most prominent when removing the dorsal cilia
in the diencephalic ventricle. All ROIs (except ROI 1, where we impose the photoconversion curve)
exhibit significant changes in the dynamics of $\cbar$
(\Cref{fig:figure5_compare_cilia_modifications}b).
Compared to the baseline model results, we observed a slower increase and lower final concentration
in the middle and anterior ventricles (ROIs 2--5),
and more rapid transport to the posterior ventricle (ROI 6).
In contrast, removal of telencephalic cilia did not alter the evolution of the
mean concentration $\cbar$ with time (\Cref{fig:figure5_compare_cilia_modifications}b).
Finally, without ventral cilia in the diencephalic ventricle,
more solute has spread towards the anterior ventricle (ROI 5) and
less solute has spread towards the posterior ventricle (ROI 6).
\begin{figure}[H]
    \centering
    \includegraphics[width=\textwidth]{graphics/figure5_compare_cilia_modifications.png}
    \caption{\textbf{a.} The three separate ciliated regions where loss of cilia motility is simulated
    by imposing a zero tangential traction boundary condition in the respective region.
    \textbf{b.} Mean concentration $\cbar(t)$ for the six regions of interest in the three 
    loss-of-cilia-motility scenarios considered. The region named in the legend is the region
    where we modify the cilia tangential traction boundary condition to a free-slip condition
    in the CSF flow model.}
    \label{fig:figure5_compare_cilia_modifications}
\end{figure}

@@

In addition to its favorable mass-conserving property, the BDM-DG scheme
lets us impose the impermeability condition strongly, and the tangential
traction boundary condition weakly, without any tailored approach,
since these boundary conditions appear naturally in the weak form of the
Stokes equations~(\cref{subsec:appendixA1}). If instead e.g. Taylor-Hood
elements were to be used, the impermeability boundary condition must
be handled with for example a Nitsche method~\cite{Nitsche1971UberSind}
or with a Lagrange multiplier~\cite{Babuska1973TheMultipliers, Bertoluzza2017BoundaryHemodynamics}.

>>>>

\documentclass[fleqn]{wlscirep}
\usepackage[utf8]{inputenc}
\usepackage{lipsum}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsfonts}
\usepackage{siunitx}
\usepackage{physics}
\usepackage{stmaryrd} % double brackets for jump operator
\usepackage{nameref}
\usepackage{booktabs}
\usepackage{bm}
\usepackage{amsopn}
\usepackage[skip=10pt]{caption}
\usepackage{color}
\usepackage{xcolor,colortbl}
\usepackage[nameinlink, capitalise]{cleveref}
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{tablefootnote} % for table footnotes
\usepackage{float} % figure placement

@@

% Title
\title{The role of advection versus diffusion in intraventricular transport}
\author[1,*]{Halvor Herlyng}
\author[1]{Ada J.~Ellingsrud}
\author[1]{Miroslav Kuchta} 
\author[\fixme{X}]{Inyoung Jeong} % OK?
\author[1,3,+]{Marie E.~Rognes} % OK order?
\author[2,+, *]{Nathalie Jurisch-Yaksi}
\affil[1]{Department of Numerical Analysis and Scientific Computing,
Simula Research Laboratory, Oslo, Norway}
\affil[2]{\fixme{@Nathalie is this correct?: Department of Clinical and Molecular Medicine,
Norwegian University of Science and Technology, Trondheim, Norway \\
Same affiliation for Inyoung?}}
\affil[3]{K.~G.~Jebsen Center for Brain Fluid Research}
\affil[*]{\href{mailto:hherlyng@simula.no}{hherlyng@simula.no}
\href{mailto:nathalie.jurisch-yaksi@ntnu.no}{nathalie.jurisch-yaksi@ntnu.no}}
\affil[+]{these authors contributed equally to this work}
% Optional PDF information
\ifpdf
\hypersetup{
  pdftitle={zebrafish-cilia-csf},
  pdfauthor={H.~Herlyng~\emph{et al.}}
}
\fi

@@

\section*{Methods}
\subsection*{Image-based computational geometries of zebrafish brain ventricles}
To model flow and transport in zebrafish larval brain ventricles, we generated a
3D representation of the brain ventricular walls using confocal imaging data of a zebrafish embryo
injected intraventricularly with a 70 kDa dye at 2 dpf~\cite{Olstad2019CiliaryDevelopment}. 
From the imaging data, we constructed a computational mesh of the ventricles (\Cref{fig:fig1}\fixme{d}).
The interior volume was meshed with fTetWild~\cite{Hu2020FastWild}. 
The \emph{standard} mesh consisted of 132 134 tetrahedral cells and had a
maximal (minimal) edge length of 14.9 $\mu$m (4.68 $\mu$m). 
At 2 dpf, the ventricular system consists mainly of 3 cavities: the telencephalic (anterior),
the di-/mesencephalic (middle) and the rhombencephalic (posterior) ventricles, which are connected by ducts. 
Parts of the ventricular surfaces were marked to be lined with
motile cilia~\cite{Olstad2019CiliaryDevelopment} (\Cref{fig:fig1}\fixme{d}, magenta markers).
\begin{figure}%[hbt!]
    \centering
    \includegraphics[width=\linewidth]{graphics/figure1.png}
    \caption{\fixme{Reorder panels.} 
    \textbf{a.} Schematic illustration of the two cerebrospinal fluid (CSF) flow components modeled,
    motile cilia and cardiac pulsations, and their contributions to the flow patterns.
    \textbf{b.} Particle tracking data visualizing flow directionality in the middle ventricle.
    \textbf{c.} Flow velocity magnitudes in the middle ventricle from particle tracking data.
    \textbf{d.} Computational mesh with marked facets indicating regions of motile cilia (magenta) and
    anterior/posterior facets (blue).
    \textbf{e.} Streamlines of the (steady-state) CSF flow field simulated with the
    cilia-only/no-cardiac flow model.
    \textbf{f.} Streamlines of the CSF flow field simulated with the cardiac-only/no-cilia flow
    model at a time 25\% into the cardiac cycle. %t=0.112613 s, period=0.45045 s
    \textbf{g.} Streamlines simulated with the baseline flow model at
    a time 50\% into the cardiac cycle. %t=0.225225 s
    \textbf{h.} Streamlines simulated with the baseline flow model at
    a time 70\% into the cardiac cycle. %t=0.315315 s
    }
    \label{fig:fig1}
\end{figure}

@@

\subsection*{Computational CSF dynamics}
The beating motion of the cilia generates CSF flow with steady rotational
structures~\cite{Olstad2019CiliaryDevelopment}
within the brain ventricles (\Cref{fig:fig1}\fixme{a--c}).
In addition, cardiac pulsations induces pulsatile CSF flow.
We model this flow of CSF in the brain ventricles by the time-dependent,
incompressible Stokes equations which read as follows.
Find the CSF velocity $\uu = \uu(\xx, t) = (u_x(\xx, t), u_y(\xx, t), u_z(\xx, t))$
and the CSF pressure $p = p(\xx, t)$ for $\xx \in \Omega$ and time $t>0$, such that
\begin{subequations}
    \begin{align}
      \rho \pdifft{\uu} - \nabla \cdot \bsig &= \mathbf{0}
      \quad \text{in } \Omega,\\
      \nabla \cdot \uu &= 0
      \quad  \text{in } \Omega,
  \end{align}
  \label{eq:stokes_eqs}% 
\end{subequations}%
where the stress tensor is defined as $\bsig(\uu, p) = 2\mu\beps(\uu) - p\mathbf{I}$
and $\beps(\uu) = \frac{1}{2}\left(\nabla \uu + (\nabla\uu)^T\right)$ is the strain-rate tensor. 
Bold-face characters denote vectors or tensors, and $\mathbf{I}$ is the identity tensor in three dimensions.
Furthermore, $\rho=1000 \ \mathrm{kg/m^3}$ is the CSF density
and $\mu=6.97\times 10^{-4} \ \mathrm{Pa\cdot s}$ is the dynamic viscosity
of the CSF~\cite{Bloomfield1998EffectsFluid}.
We introduce the tangential traction $\bsigpar$ as the
tangential component of the traction $\hat{\bsig}=\bsig\nn$:
\begin{equation*}
    \bsigpar = P_{\nn}(\hat{\bsig}) = (\mathbf{I} - \nn\otimes\nn)\hat{\bsig},
\end{equation*}
where $\nn$ is the outer unit normal of the boundary surface $\Gamma = \partial\Omega$,
and $P_{\nn}(\rr)$ is the tangential projection of a vector $\rr$ onto $\Gamma$.
We consider three types of boundary conditions:
(i) \emph{tangential traction} (no flow normal to the boundary
$\uu \cdot \nn = 0$ with a force $\btau$ applied tangentially on the boundary $\bsigpar = \btau$),
(ii) \emph{normal pressure} (a pressure $\Tilde{p}$ applied in
the normal direction: $(\mu\nabla\uu - p\mathbf{I})\nn = \Tilde{p}(t)\nn$),
and (iii) \emph{free slip} (no flow normal to the boundary
$\uu \cdot \nn = 0$ and no tangential forces $\bsigpar = {\bf 0}$).
Unless noted otherwise, the free-slip condition is applied on the
boundary. We assume that the system starts at rest: $\uu(\xx, t=0) = \mathbf{0}$.

@@

\subsection*{Cilia-driven flow}
On the ciliated regions of the ventricle walls $\Gc$ (\Cref{fig:fig1}\fixme{d}, magenta markers),
we impose a constant-in-time tangential traction $\bsigpar=\btau$ to represent the net forces
of the cilia acting on the CSF.
We set $\btau(\xx, \rr) = \tau \lambda(\xx) P_{\nn}(\rr)$, where the function $\lambda(\xx)$
and the sign of the vector $\rr=\pm(1, 0, 1)$ varies depending on the
cilia region (see~\Cref{subsec:appendixA1}).
Both $\rr$ and the value $\tau=6.5\times 10^{-4}$ Pa were chosen by 
model calibration using experimental data~\cite{Olstad2019CiliaryDevelopment}.
The data indicate speeds proximate to the ventricular walls of $27.4 \pm 5.4$ \textmu m/s and
$2.6 \pm 0.6$ \textmu m/s in the dorsal and ventral regions of the middle ventricle, respectively.
In the anterior ventricle, the speeds were estimated at $4.7 \pm 1.3$ \textmu m/s.
We calibrated $\tau$ heuristically by comparing these experimental velocity values
with the simulated, maximum velocity magnitude.

\subsection*{Flow induced by cardiac pulsations}\label{subsec:pressure_bcs}
To drive a cardiac-induced pulsatile flow inside the brain ventricles, we set $\Tilde{p}(t)=0$ and
$\Tilde{p}(t)=-A\sin{\omega t}$ on anterior and posterior parts, respectively, of the
boundary $\Gp$~(\Cref{fig:fig1}\fixme{d}, blue markers).
Based on a measured cardiac frequency of $f=2.22$ Hz~\cite{Olstad2019CiliaryDevelopment},
the angular cardiac frequency $\omega=2\pi f=6.97$ rad/s.
The amplitude $A=1.5\times 10^{-3}$ Pa determines the magnitude of the pulsatile flow,
and was calibrated based on experimental data~\cite{Olstad2019CiliaryDevelopment}.
Similar to $\tau$, the amplitude $A$ was chosen by comparing experimental and simulated velocities.
Particle tracking data of pulsatile motion was used to determine the contributions of the
cardiac pulsations to the velocity magnitude,
as compared to the effect of cilia motion on the flow (see~\Cref{sec:appC_particle_transport}).

@@

\subsection*{Intraventricular injection of microbeads and particle tracking}
To recover the flow fields shown in~\Cref{fig:fig1}\fixme{b, c}, we carried out
particle tracking in a larval zebrafish injected with fluorescent beads.
As described in Olstad~\emph{et al.}~\cite{Olstad2019CiliaryDevelopment},
anesthetized 2 dpf zebrafish were injected with 1 nl of a mixture containing 0.1\% w/v fluorescent beads
(SPHERO Fluorescent Yellow Particles 1\% w/v, F = 0.16 mm) diluted
in 7.5 mg/ml 70 kDa rhodamine B isothiocyanate-dextran (RITCdextran; Sigma-Aldrich, R9379)
dissolved in artificial CSF. Artificial CSF composition was as follows:
124 mM NaCl, 22 mM D-(+)-Glucose, 2.0 mM KCl, 1.6 mM MgSO\textsubscript{4} $\cdot$ 7 H\textsubscript{2}O,
1.3 mM KH\textsubscript{2}PO\textsubscript{4}, 24 mM NaHCO\textsubscript{3},
2.0 mM CaCl\textsubscript{2} $\cdot$ 2 H\textsubscript{2}O.
The needles used for the injections were pulled with a
Sutter Instrument Co. Model P-2000 from thin-walled glass
capillaries (1.00 mm; VWR) and cut open with a forceps.
A volume of 1 nl of the solution was injected with a pressure injector
(Eppendorf Femtojet 4i) in the rostral rhombencephalic ventricle.
The pressure and time were calibrated for each needle using a 0.01 mm calibration slide.

Following injection, the zebrafish was directly placed under the confocal
microscope and 1200 images were acquired at a frequency of 13.1 Hz.
A single optical section was obtained. Particle tracking was done using
TrackMate~\cite{Tinevez2017TrackMate:Tracking} in Fiji/ImageJ~\cite{Schindelin2012Fiji:Analysis}
and plotted using MATLAB, as described in Jeong~\emph{et al.}~\cite{Jeong2022MeasurementTelencephalon}.
The parameters for TrackMate were as follows.
The DoG detector was used with the settings: threshold: 2.0, with
median filtering; radius: 2.0, with subpixel localization.
Next, the Simple LAP tracker was used with settings: max frame gap: 2;
alternative linking cost factor: 1.05; linking max distance: 5.0; gap closing max distance: 2.0;
splitting max distance: 15.0; allow gap closing: true; allow track splitting:
false; allow track merging: false; merging max distance: 15.0; cutoff percentile: 0.9.
Only the particle tracks with at least 30 data points, while simultaneously
covering a distance of at least 8.8 $\mu$m, were plotted in~\Cref{fig:fig1}\fixme{b, c}.

@@

We seek to model the compartimentalized and pulsatile CSF flow observed \emph{in vivo}~(\Cref{fig:fig1}\fixme{b--c}).
To characterize CSF flow patterns predicted by
computational fluid dynamics within the brain ventricles, we begin by
considering the simulated CSF velocity and pressure when only considering
cilia motion, disregarding cardiac pressure pulsations. The
bidirectional traction induced by the motile cilia
inherently leads to a partial compartmentalization of the ventricular
system, with large-scale vortex structures in the anterior and middle
ventricles~(\Cref{fig:fig1}\fixme{e}). The highest velocities ($26.5$ \textmu m/s)
occur in the vicinity of the dorsal cilia in the middle ventricle.

To examine the effect of cardiac pressure pulsations,
we next consider a no-cilia model variation. In this scenario, the vortex structures
generated by the ciliary motion vanish. The simulated CSF
flow is pulsatile as a result of the sinusoidal pressure gradient (\Cref{fig:fig1}\fixme{g}),
with a peak velocity magnitude of $4.8$ \textmu m/s. At each time instant, the flow
is unidirectional, with CSF initially flowing rostrocaudally (positive $x-$axis) before changing
direction to caudorostral in the middle of the cardiac cycle.

Next we consider the baseline model, simulating flow driven by both cilia
and cardiac pulsations. We observe persistent flow features,
with vortex-structures in the anterior and middle ventricles
and directional flow in the posterior ventricle (\Cref{fig:fig1}\fixme{e},
\fixme{Supplementary video 1}). These characteristic patterns suggest
that the flow in the anterior and middle ventricles are dominated by
the cilia contribution, while the cardiac pressure pulsations dominate
within the posterior ventricle. Over one cardiac cycle, the CSF flow
speed reaches up to 27.7 \textmu m/s, dorsally in the posterior part
of the middle ventricle. This peak is attained at the time instant
coinciding with the time of peak caudal cardiac pulsatile flow, as a
result of both the traction induced by the cilia and the pulsatile
pressure-driven flow being aligned with the rostrocaudal 
axis. Using the observed maximum velocity magnitude and choosing the height of
the middle ventricle ($110$ \textmu m) as a characteristic length
scale, yields a Reynolds number $\mathrm{Re} \approx 0.004$,
justifying the assumption of Stokes flow. In summary, the computational
CSF flow model reproduces the flow features observed \emph{in vivo}.

@@

\subsection*{Absence of ciliary motion affects local protein distribution}
Since CSF flow is highly disturbed in the absence of motile cilia, we
want to assess the impact of cilia motion on solute distribution. To this end,
we measured the movement of photoconverted Dendra2 signal
in control and motile-cilia mutant zebrafish (\emph{schmalhans} (\emph{smh})).
By plotting the means of $\Delta F$ (\Cref{fig:figure4_compare_control_mutant}a)
and the time to threshold
across the control and mutant cohorts (\Cref{fig:figure4_compare_control_mutant}b),
we identified a significant
difference in Dendra2 distribution in ROIs 2 and 4, but not in ROIs 3, 5, and 6. In ROIs 2 and 4,
which measure Dendra2 distribution downstream of the photoconversion site,
we observed a slower increase of fluorescence intensity,
resulting in $\sim 45\%$ (ROI 2) and \fixme{$\sim X\%$ (ROI 4)}
higher mean times-to-threshold in the \emph{smh}
mutants compared to controls (\Cref{fig:figure4_compare_control_mutant}b).

@@

We identified that the velocities were most strongly
affected when removing cilia in the dorsal diencephalic cilia,
as the maximum velocity dropped from 27.7 $\mu$m/s to 8.0 $\mu$m/s.
\emph{In vivo}, these cilia generate the strongest flow at
this developmental stage of the fish~\cite{Olstad2019CiliaryDevelopment}.
In contrast, removing telencephalic and ventral diencephalic had minimal
impact and resulted in maximum velocities of 27.7 $\mu$m/s or 27.2 $\mu$m/s, respectively.
Changes in the transport dynamics of Dendra2 were most prominent when removing the dorsal cilia
in the diencephalic ventricle. All ROIs (except ROI 1, where we impose the photoconversion curve)
exhibit significant changes in the dynamics of $\cbar$
(\Cref{fig:figure5_compare_cilia_modifications}b).
Compared to the baseline model results, we observed a slower increase and lower final concentration
in the middle and anterior ventricles (ROIs 2--5),
and more rapid transport to the posterior ventricle (ROI 6).
On the other hand, removal of telencephalic cilia did not alter the evolution of the
mean concentration $\cbar$ with time (\Cref{fig:figure5_compare_cilia_modifications}b).
Finally, without ventral cilia in the diencephalic ventricle,
more solute has spread towards the anterior ventricle (ROI 5) and
less solute has spread towards the posterior ventricle (ROI 6).
\begin{figure}[H]
    \centering
    \includegraphics[width=\textwidth]{graphics/figure5_compare_cilia_modifications.png}
    \caption{\textbf{a.} The three separate ciliated regions where loss of cilia motility is simulated
    by imposing a zero tangential traction boundary condition in the respective region.
    \textbf{b.} Mean concentration $\cbar(t)$ for the six regions of interest in the three 
    loss-of-cilia-motility scenarios considered. The region named in the legend is the region
    where we modify the cilia tangential traction boundary condition to a free-slip condition
    in the CSF flow model.}
    \label{fig:figure5_compare_cilia_modifications}
\end{figure}

@@

In addition to its favorable mass-conserving property, the BDM-DG scheme
lets us impose the impermeability condition strongly, and the tangential
traction boundary condition weakly, without any tailored approach,
since these boundary conditions appear naturally in the weak form of the
Stokes equations~(\Cref{subsec:appendixA1}). If instead e.g. Taylor-Hood
elements were to be used, the impermeability boundary condition must
be handled with for example a Nitsche method~\cite{Nitsche1971UberSind}
or with a Lagrange multiplier~\cite{Babuska1973TheMultipliers, Bertoluzza2017BoundaryHemodynamics}.
